---
title: "Phenology KPI"
author: "Katie Jones"
date: "Thursday, June 04, 2015"
output: word_document
---

```{r libraries}

library(plyr)
library(dplyr)
library(atbdLibrary)
library(xtable)

```



```{r readNPNdata}
#set working directory
setwd("~/R/phenology/NPNdata_20150715")


allData <- read.csv('observation_data.csv', na.strings = "",
                    stringsAsFactors=F)

```


```{r manipulate dataset}

#subset to to phenophase status='yes' (1), and reduce # of fields.
df1 <- filter(allData, Phenophase_Status==1) %>%
  select(Observation_Date, ObservedBy_Person_ID, Individual_ID,
         USDA_PLANTS_Symbol, Species_Functional_Type, Phenophase_Name,
         Phenophase_Status, Intensity_Value, Genus, Species) 


#create scientificName field
df1$scientificName <- "na"
df1$scientificName <- paste(df1$Genus, df1$Species, sep=" ")


#remove old dfs
#rm(allData)

```


```{r, reduce nrows to just dupes}
#Set temp column in order to identify records with duplicate values in specific fields
colsToValidate <- c('Individual_ID', 'Observation_Date', 'Phenophase_Name')

#select only those row that have duplicate values
df2<-df1[duplicated (df1[,colsToValidate]) | duplicated (df1[,colsToValidate], fromLast=TRUE),]

# remove duplicates resulting from multiple observations recorded by only one person on a given day
df2 <- rem_dups(df=df2, cols=c(colsToValidate, 'ObservedBy_Person_ID'), qfColName = 'dupByPerson')

# remove phenophases for which no intensity is reported
df2 <- filter(df2, !Phenophase_Name=="Falling leaves", !Phenophase_Name=="Pollen release",
              !Phenophase_Name=="Flower heads", !Phenophase_Name=='Flowers',
              !Phenophase_Name=='Flowers or flower buds', !Phenophase_Name=='Open pollen cones', 
               !Phenophase_Name=='Pollen cones', !Intensity_Value==-9999)


df2$temp <- paste(df2$Observation_Date, df2$Individual_ID, df2$Phenophase_Name, 
                   sep='_')

#Set empty vectors to hold the values needed for the new dataframe
dates <- vector()
indIDs <- vector()
pheNames <- vector()
agrees <- vector()
numObs <- vector()
funcType <- vector()


#Subset data by unique values of temp
for (i in 1:length(unique(df2$temp))){
  sub <- df2[which(df2$Observation_Date==strsplit(df2$temp[i],'_')[[1]][1] &
                     df2$Individual_ID==strsplit(df2$temp[i],'_')[[1]][2] & 
                     df2$Phenophase_Name==strsplit(df2$temp[i],'_')[[1]][3]),]
  

  #If there is more than one row in the subset, calculate agree (0=no, 1=yes)
  if(nrow(sub)>1){
    agree <- ifelse(length(unique(sub$Intensity_Value))==1, 1, 0)
    numObs <- c(numObs,nrow(sub))
    dates <- c(dates,sub$Observation_Date[1])
    indIDs <- c(indIDs,sub$Individual_ID[1])
    pheNames <- c(pheNames,sub$Phenophase_Name[1])
    funcType <- c(funcType, sub$Species_Functional_Type [1])
    agrees <- c(agrees,agree)
  }
}

#Create output dataframe
out <- data.frame(date=dates, individualID=indIDs, phenophaseName= pheNames,
                  functionalType=funcType, agree=agrees,
                  numberObservers=numObs)

out <- rem_dups(out, names(out))

```


```{r summary}


## overall concurance of phenophase intensity for = 0.8769671
propConcur <- (sum(out$agree))/(nrow(out))
propConcur


#summerize by phenophase Name only
propSummaryPhe <- out %>% 
  group_by(phenophaseName) %>%
  summarise(sumAgree=sum(agree), sumDist=sum(dist),  instances=length(agree), propAgree=sumAgree/instances, propDist=sumDist/instances)

propSummaryPhe2

#summerize by functional type
propSummaryType <- out %>% 
  group_by(functionalType) %>%
  summarise(sumAgree=sum(agree), instances=length(agree), propAgree=sumAgree/instances)

propSummaryType

#remove functional types with n<100
propSummaryType <- filter(propSummaryType, !functionalType=="Graminoid")
propSummaryType <- filter(propSummaryType, !functionalType=="Cactus")
propSummaryType <- filter(propSummaryType, !functionalType=="Forb")

out2 <- filter(out, !functionalType=="Graminoid", !functionalType=="Cactus", !functionalType=="Forb")

propConcur2 <- (sum(out2$agree))/(nrow(out2))
propConcur2

propSummaryType2 <- out2 %>% 
  group_by(functionalType) %>%
  summarise(sumAgree=sum(agree), instances=length(agree), propAgree=sumAgree/instances)

propSummaryType2

```


```{r printable tables}

propSumPhe.Table <- xtable(propSummaryPhe)
print.xtable(propSumPhe.Table, type="html", file="noskill_propSumPhe.html")

propSummaryType.Table <- xtable(propSummaryType)
print.xtable(propSummaryType.Table, type="html", file="noskill_propSummaryType.html")

propSummaryType2.Table <- xtable(propSummaryType2)
print.xtable(propSummaryType2.Table, type="html", file="noskill_propSummaryType2.html")

```



```{r reassign intensity}

df3 <- df2

unique(df3$Intensity_Value)
unique(df3$Phenophase_Name)

df3$Intensity_Value <- as.character(df3$Intensity_Value)
df3$Intensity_Value[df3$Intensity_Value == "Less than 3"] <- 1
df3$Intensity_Value[df3$Intensity_Value == "3 to 10"] <- 2
df3$Intensity_Value[df3$Intensity_Value == "11 to 100"] <- 3
df3$Intensity_Value[df3$Intensity_Value == "101 to 1,000"] <- 4
df3$Intensity_Value[df3$Intensity_Value == "1,001 to 10,000"] <- 5
df3$Intensity_Value[df3$Intensity_Value == "More than 10,000"] <- 6
df3$Intensity_Value[df3$Intensity_Value == "Less than 5%"] <- 7
df3$Intensity_Value[df3$Intensity_Value == "5-24%"] <- 8
df3$Intensity_Value[df3$Intensity_Value == "Less than 25%"] <- 8
df3$Intensity_Value[df3$Intensity_Value == "25-49%"] <- 9
df3$Intensity_Value[df3$Intensity_Value == "50-74%"] <- 10
df3$Intensity_Value[df3$Intensity_Value == "75-94%"] <- 11
df3$Intensity_Value[df3$Intensity_Value == "95% or more"] <- 12

df3$Intensity_Value <- as.integer(df3$Intensity_Value)
df3 <- filter(df3, !Intensity_Value=='NA')


##max(Intensity_Value)-min(Intensity_Value)

#Set empty vectors to hold the values needed for the new dataframe
dates <- vector()
indIDs <- vector()
pheNames <- vector()
agrees <- vector()
numObs <- vector()
funcType <- vector()
minMaxDistance <- vector()


#Subset data by unique values of temp
for (i in 1:length(unique(df3$temp))){
  sub <- df3[which(df3$Observation_Date==strsplit(df3$temp[i],'_')[[1]][1] &
                     df3$Individual_ID==strsplit(df3$temp[i],'_')[[1]][2] & 
                     df3$Phenophase_Name==strsplit(df3$temp[i],'_')[[1]][3]),]
  

  #If there is more than one row in the subset, calculate agree (0=no, 1=yes)
  # calculate intensity value distance between disagreement
  if(nrow(sub)>1){
    agree <- ifelse(length(unique(sub$Intensity_Value))==1, 1, 0)
    numObs <- c(numObs,nrow(sub))
    diff <- max(sub$Intensity_Value)-min(sub$Intensity_Value)
    dates <- c(dates,sub$Observation_Date[1])
    indIDs <- c(indIDs,sub$Individual_ID[1])
    pheNames <- c(pheNames,sub$Phenophase_Name[1])
    funcType <- c(funcType, sub$Species_Functional_Type [1])
    agrees <- c(agrees,agree)
    minMaxDistance <- c(minMaxDistance, diff)
  }
}


outDist <- data.frame(date=dates, individualID=indIDs, phenophaseName= pheNames,
                  functionalType=funcType, agree=agrees,
                  numberObservers=numObs, minMaxDistance=minMaxDistance)

outDist <- rem_dups(outDist, names(outDist))
outDist$dist <- 'na'

outDist$dist <- ifelse((outDist$minMaxDistance==0 | outDist$minMaxDistance==1), 
                       outDist$dist <- 1,outDist$dist <- 0)

outDist$intensityType <- 'NA'
outDist$intensityType[outDist$phenophaseName == "Breaking leaf buds"] <- 'count'
outDist$intensityType[outDist$phenophaseName == "Emerging needles"] <- 'count'
outDist$intensityType[outDist$phenophaseName == "Young leaves"] <- 'count'
outDist$intensityType[outDist$phenophaseName == "Young needles"] <- 'count'
outDist$intensityType[outDist$phenophaseName == "Colored leaves"] <- 'percentage'
outDist$intensityType[outDist$phenophaseName == "Increasing leaf size"] <- 'percentage'
outDist$intensityType[outDist$phenophaseName == "Leaves"] <- 'percentage'
outDist$intensityType[outDist$phenophaseName == "Open flowers"] <- 'percentage'



propConcurDist <- (sum(outDist$dist))/(nrow(outDist))
propConcurDist
 
propSummaryTypeDist <- outDist %>% 
  group_by(functionalType) %>%
  summarise(sumDist=sum(dist), instances=length(dist), propAgree=sumDist/instances)

propSummaryTypeDist

median(propSummaryTypeDist$propAgree)

propSummaryPheDist <- outDist %>% 
  group_by(phenophaseName) %>%
  summarise(sumDist=sum(dist), instances=length(dist), propAgree=sumDist/instances)

propSummaryPheDist

median(propSummaryPheDist$propAgree)


propSummaryIntensityType <- outDist %>% 
  group_by(functionalType, intensityType) %>%
  summarise(sumAgree=sum(agree), instances=length(agree), propAgree=sumAgree/instances)

propSummaryIntensityType

propSummaryType2 <- outDist %>% 
  group_by(functionalType) %>%
  summarise(sumAgree=sum(agree), sumDist=sum(dist),  instances=length(agree), propAgree=sumAgree/instances, propDist=sumDist/instances)

propSummaryType2

propSummaryType2 <- filter(propSummaryType2, !functionalType=="Cactus")
propSummaryType2 <- filter(propSummaryType2, !functionalType=="Forb")

propSummaryTypeInt <- outDist %>% 
  group_by(functionalType, intensityType) %>%
  summarise(sumAgree=sum(agree), sumDist=sum(dist),  instances=length(agree),
            propAgree=sumAgree/instances, propDist=sumDist/instances)

propSummaryTypeInt

propSummaryTypeInt <- filter(propSummaryType2, !functionalType=="Cactus")
propSummaryTypeInt <- filter(propSummaryType2, !functionalType=="Forb")

propSummaryTypeInt

#Print tables
propSumPheDist.Table <- xtable(propSummaryPheDist)
print.xtable(propSumPheDist.Table, type="html", file="noskill_propSumPheDist.html")

propSummaryTypeDist.Table <- xtable(propSummaryTypeDist)
print.xtable(propSummaryTypeDist.Table, type="html", file="noskill_propSummaryTypeDist.html")

propSummaryIntensityType.Table <- xtable(propSummaryIntensityType)
print.xtable(propSummaryIntensityType.Table, type="html", file="propSummaryIntensityType.html")

```



```{r anova}
#does experience matter?

skill <- lm(agree~skillLevel, data=out)
anova(skill)

type <- lm(agree~functionalType, data=out)
anova(type)
  #stongly influenced by ddb, test without
out2 <- filter(out, !functionalType=='Deciduous broadleaf')
omit_ddb <- lm(agree~functionalType, data=out2)
anova(omit_ddb)


t <- lm(agree~skillLevel+phenophaseName+skillLevel*phenophaseName, data=out)
anova(t)

plot(propagree ~ phenophaseName, data = propSummarySkillPhe)
         

```


We looked across growth forms 
median proportion of agreement assuming <100 found it be %73. We feel this is low given the attention 
to training and experience of NEON crew, bump up to %85. NPN data are reported by observers with highly variable experience 
Citizen Sci
dumped a lot of data...


categorize intensity values by percentage vs counts

at the point of initial commissioning agreement must be within the growthform levels, additional assessment should demonstrated

Counts = same bin (log scale)
percentage = within one

Can we rank? use agreement WITHIN bins as absolute fail threshold but 'quality' would be at a higher level. Ask Greg about grading.
